Fixing Vulnerabilities Potentially Hinders Maintainability

### 2020-05-25 paper/icsme20.tex
small update
new comments adressed + luis comments added

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{framed}
\usepackage{float}
\usepackage{xcolor}
\usepackage{footnote}
\usepackage{xfrac}
\usepackage{balance}
\usepackage{hyperref}
\usepackage{array}
\usepackage{subfigure}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

% \title{Fixing Vulnerabilities Hinders Maintainability}
\title{\Sof{TBA - New Titles in the comments}}
%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
%\title{The reason why you should be applying coding practices: safer software.}
%\title{What is the impact of security patches in software evolution?}
%\title{Hey, developers! You should pay more attention to your patches.}
%\title{Are good coding pratices used when patching software vulnerabilities?}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of 10 
guidelines using a dataset including 1335 security commits.
Results show that fixing software vulnerabilities may hinder the maintainability 
of open-source software and that different levels of attention should be given to 
different severity levels, programming languages and types of weaknesses. We 
conclude that changes to code bases to patch 
vulnerabilities need to be performed with extra care and that security-related 
code evolution should be subject in computer curricula.
\end{abstract}

@@

\textbf{Guidelines/Metrics:} Each patch performs a set of changes
on the source code of the software. These changes may have a different
impact on the guidelines/metrics used to measure software maintainability. 
Figure~\ref{fig:guidelines} shows the impact of security patches 
on each guideline individually and the average impact on all guidelines 
together - $M(v)$. \Rui{doesn't parse -->} Under each guideline is stated the 
metric used to the calculations. Table~\ref{tab:guidelines} describes 
in more detail the metrics behind the guidelines. 
For each type of guideline, three types of impact of patches in 
the software maintainability are presented: 
\emph{negative}, maintainability decreases (\emph{red});
\emph{none}, maintainability remains the same (\emph{yellow}); and,
\emph{positive}, maintainability increases (\emph{green}). 
For each type of guideline, a boxplot 
and a swarmplot are presented to show the variability and dipersion of
the results alongside the number of absolute
and relative cases of each impact. Next to each
type of guideline, it is presented the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the
Paired Wilcoxon signed-rank test. $M(v)$ is not a guideline but rather the 
average impact of all guidelines.

@@

\textbf{Overall Score ($M(v)$):} 
Although overall patching vulnerabilities has a less negative impact
on software maintainability guidelines, this is not reflected in the average 
impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
The $M(v)$ plots show that $406$ ($41.9\%$) of the cases have negative impact
on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7$)
have a negative impact. The larger number of negative cases may be 
explained by guidelines with higher concentration of negative cases with higher amplitudes
such as \Rui{these acronyms came out of the blue.} WShortUC, WSimpleUC and SCM. In other words, it is 
important to note that the design debt of one guideline can lead
to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
ESLint, Infer, and more) should be leveraged to 
assess the risk of patching vulnerabilities - alongside with pull
requests and code reviews. While static code analysis
may be daunting due to the number of rules and different effects
on maintainability, BCH only uses the top-$10$ of the guidelines
with the highest effect on maintainability. Static analysis 
can be used to complement the source code analysis, introduce
the capability of vulnerability detection, and, consequently,
reduce the risk of hindering maintainability.

 \begin{figure}[H]
  	\centering
  	\includegraphics[width=0.5\textwidth]{figures/main_per_guideline.pdf}
 	\vspace{-0.8 cm}
  	\caption{Impact of the security patches per guideline.}
 	\label{fig:guidelines}
 	\vspace{0.2 cm}
 	\scalebox{0.5}{
 	\begin{tabular}{r@{: }l r@{: }l}
 	\textbf{WCC} & Write Clean Code & \textbf{KACB} & Keep Architecture Components Balanced\\
 	\textbf{CACL} & Couple Architecture Components Loosely & \textbf{SCM} & Separate Concerns
	in Modules \\
 	\textbf{KUIS} & Keep Unit Interfaces Small & \textbf{WCO} & Write Code Once \\
 	\textbf{WSimpleUC} & Write Simple Units of Code & \textbf{WShortUC} & Write Short Units of
	Code
 	\end{tabular}
 	}
 	\vspace{-0.3 cm}
 \end{figure}

\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
Database} (NVD) website to obtain their severity levels. In total, we 
retrieved severity scores for $536$ vulnerabilities: $112$ (\emph{HIGH}), 
$395$ (\emph{MEDIUM}) and $29$ (\emph{LOW}). Figure~\ref{fig:severity} 
presents the impact of security patches per severity level on the 
maintainability of open-source software. We observe that patches for 
\emph{HIGH} ($50.0\%$) and \emph{MEDIUM} ($43.8\%$) severity vulnerabilities 
hinder more the maintainability of software than \emph{LOW} ($27.6\%$) 
severity vulnerabilities. Again, patches have a considerable 
negative impact on software maintainability - between $20\%$ 
and $50\%$. Statistical significance was retrieved only for \emph{LOW}
severity vulnerabilitie, i.e., \emph{LOW} severity vulnerabilities
may have less negative impact on software maintainability thant the 
other severity levels. However, 
results should not be disregarded because they seem to confirm the 
assumption that higher severity vulnerabilities patches may have a 
more negative impact on maintainability and, consequently, that high/medium 
severity vulnerabilities may need more attention than low severity while 
patching vulnerabilities.

@@

\textbf{Programming Languages:} The impact on software maintainability per programming 
language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
this analysis to programming languages with at least $20$ data points, as this 
is a requirement for our hypothesis tests (HT). Thus, we compare the results for 
\texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
\texttt{Python} and \texttt{PHP}. \texttt{Groovy} samples are not presented
because the number of cases did not fulfill the HT requirement. 
\texttt{C/C++} ($46.5\%$), \texttt{Ruby} ($46.5\%$) and 
\texttt{PHP} ($38.6\%$) are the programming languages with worse 
impact on software maintainability. \texttt{Java} ($50.9\%$) and \texttt{Python} 
($46.5\%$) seem to be the 
languages less affected by patching, i.e., more than $45\%$ of patches had a positive 
impact on maintainability. But overall patches in all languages have a 
considerable negative impact on maintainability - between $35\%$ to $50\%$ - which 
confirms the need for better and more secure programming languages. 
Statistical significance was only retrieved for the \texttt{C/C++} 
(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} (p-value = $0.048$) languages.
Results should be read with care where statistical significance was not found.
Yet, data reports very interesting hints on the impact of programming languages 
on security patches.

@@

\textbf{Conclusion:} Results show that developers may fail to follow the coding
practice guidelines and hinder software maintainability after patching; and, 
that different levels of attention should be paid to each guideline. For instance,
\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
are the most affected ones. No statistical significance was retrieved
for \emph{Separate Concerns in Modules}. Different severity vulnerabilities
may need different levels of attention and better and more secure programming languages
are needed.
The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
is $0.044$. Since the p-value is 
below the significance level of $0.05$, we argue that \emph{security patches 
may have a negative impact on the maintainability of open-source software}; and, 
that tools such as Better Code Hub should be integrated in the CI/CD pipeline
to help developers evaluate the risk of patches hindering software maintainability
alongside with Pull Requests/Code Reviews.

@@

@@

For both types of changes, there is a negative impact
on software maintainability - especially, for security changes. This raises 
a new trade-off when developers need to patch vulnerabilities on their projects
because they might be hindering the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to patch and produce secure code without hindering the maintainability of
their projects. But there are still some concerns that need to be addressed
and that this study tries to bring awareness for:

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{framed}
\usepackage{float}
\usepackage{xcolor}
\usepackage{footnote}
\usepackage{xfrac}
\usepackage{balance}
\usepackage{hyperref}
\usepackage{array}
\usepackage{subfigure}
\usepackage{changepage}

\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\title{Fixing Vulnerabilities Potentially Hinders Maintainability}
%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
%\title{The reason why you should be applying coding practices: safer software.}
%\title{What is the impact of  corollarysecurity patches in software evolution?}
%\title{Hey, developers! You should pay more attention to your patches.}
%\title{Are good coding pratices used when patching software vulnerabilities?}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of 10 
guidelines using a dataset including 1335 security commits.
Results show that fixing software vulnerabilities may hinder the maintainability 
of open-source software and that different levels of attention should be given to 
different severity levels, programming languages and types of weaknesses. We 
conclude that changes to code bases to patch 
vulnerabilities need to be performed with extra care and that security-related 
code evolution should be subject in computer curricula.\Rui{Needs improvement}
\end{abstract}

@@

\textbf{Guidelines/Metrics:} Each patch performs a set of changes
on the source code of the software. These changes may have a different
impact on the guidelines/metrics used to measure software maintainability. 
Figure~\ref{fig:guidelines} shows the impact of security patches 
on each guideline individually and the average impact on all guidelines 
together - $M(v)$. \Rui{doesn't parse -->} Under each guideline, it is stated the 
metric used to the calculations. Table~\ref{tab:guidelines} describes 
in more detail the metrics behind the guidelines. 
For each type of guideline, a boxplot 
and a swarmplot are presented to show the variability and dipersion of
the results alongside the number of absolute
and relative cases of each impact. Next to each
type of guideline, it is presented the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the
Paired Wilcoxon signed-rank test. $M(v)$ is not a guideline but rather the 
average impact of all guidelines.

@@

\textbf{Overall Score ($M(v)$):} 
Although overall patching vulnerabilities has a less negative impact
on software maintainability guidelines, this is not reflected in the average 
impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
The $M(v)$ plots show that $406$ ($41.9\%$) cases have negative impact
on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7\%$) cases
have a negative impact. The larger number of negative cases may be 
explained by guidelines with higher concentration of negative cases with higher amplitudes, 
such as, \emph{Write Short Units of 
Code}, \emph{Write Simple Units of 
Code} and \emph{Separate Concerns in Modules}.

% For each type of guideline, three types of impact of patches in
% the software maintainability are presented:
% \emph{negative}, maintainability decreases (\emph{red});
% \emph{none}, maintainability remains the same (\emph{yellow}); and,
% \emph{positive}, maintainability increases (\emph{green}).

 \begin{figure}[htp]
     \begin{adjustwidth}{-2.5cm}{-1cm}

  	\centering
  	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
  	\caption{Impact of the security patches per guideline.}
 	\label{fig:guidelines}
    \end{adjustwidth}
 	\vspace{-0.5 cm}

\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
Database} (NVD) website to collect their severity levels. In total, we 
retrieved severity scores for $536$ vulnerabilities: $112$ (\emph{HIGH}), 
$395$ (\emph{MEDIUM}) and $29$ (\emph{LOW}). Figure~\ref{fig:severity} 
presents the impact of security patches per severity level on the 
maintainability of open-source software. We observe that patches for 
\emph{HIGH} ($50.0\%$) and \emph{MEDIUM} ($43.8\%$) severity vulnerabilities 
hinder more the maintainability of software than \emph{LOW} ($27.6\%$) 
severity vulnerabilities. Again, patches have a considerable 
negative impact on software maintainability - between $20\%$ 
and $50\%$. Statistical significance was retrieved only for \emph{LOW}
severity vulnerabilitie, i.e., \emph{LOW} severity vulnerabilities
may have less negative impact on software maintainability thant the 
other severity levels. However, 
results should not be disregarded because they seem to confirm the 
assumption that higher severity vulnerabilities patches may have a 
more negative impact on maintainability and, consequently, that high/medium 
severity vulnerabilities may need more attention than low severity while 
patching vulnerabilities.

@@

\textbf{Programming Languages:} The impact on software maintainability per programming 
language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
this analysis to programming languages with at least $20$ data points, as this 
is a requirement for our hypothesis tests. Thus, we compare the results for 
\texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
\texttt{Python} and \texttt{PHP}. \texttt{Groovy} samples are not presented
because the number of cases did not fulfill the statistical test requirement. 
\texttt{C/C++} ($46.5\%$), \texttt{Ruby} ($46.5\%$) and 
\texttt{PHP} ($38.6\%$) are the programming languages with worse 
impact on software maintainability. \texttt{Java} ($50.9\%$) and \texttt{Python} 
($46.5\%$) seem to be the 
languages less affected by patching, i.e., more than $45\%$ of patches had a positive 
impact on maintainability. But overall patches in all languages have a 
considerable negative impact on maintainability - between $35\%$ to $50\%$ - which 
confirms the need for better and more secure programming languages. 
Statistical significance was only retrieved for the \texttt{C/C++} 
(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} (p-value = $0.048$) languages.
The remaining programming languages did not yield statistical significance, 
thus results should be read with care.
Yet, data reports very interesting hints on the impact of programming languages 
on security patches.

@@

\textbf{Conclusion:} Results show that developers fail to follow the coding
practice guidelines and hinder software maintainability after patching; and, 
that different levels of attention should be paid to each guideline. For instance,
\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
are the most affected ones. No statistical significance was retrieved
for \emph{Separate Concerns in Modules}. Different severity vulnerabilities
may need different levels of attention and better and more secure programming languages
are needed.
The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
is $0.044$. Since the p-value is 
below the significance level of $0.05$, we argue that \emph{security patches 
may have a negative impact on the maintainability of open-source software}; and, 
that tools such as Better Code Hub should be integrated in the CI/CD pipeline
to help developers evaluate the risk of patches hindering software maintainability
alongside with Pull Requests/Code Reviews.

@@

\Luis{
Can we say somehting on why some vulbnerabilities are more affect than others? Complexity? Hard to understand? Good  Documentation resources?
this is valid for RQ1 as well. What are the implciations of results in these different vulneratbilities?}

@@

For both types of changes, there is a negative impact
on software maintainability $-$ especially, for security changes. This raises 
a new trade-off when developers need to patch vulnerabilities on their projects
because they might be hindering the software maintainability or even introducing
new vulnerabilities.

\section{Study Implications}\label{sec:implications}

 In other words, it is 
important to note that the design debt of one guideline can lead
to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
ESLint, Infer, and more) should be leveraged to 
assess the risk of patching vulnerabilities - alongside with pull
requests and code reviews. While static code analysis
may be daunting due to the number of rules and different effects
on maintainability, BCH only uses the top-$10$ of the guidelines
with the highest effect on maintainability. Static analysis 
can be used to complement the source code analysis, introduce
the capability of vulnerability detection, and, consequently,
reduce the risk of hindering maintainability.

This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to patch and produce secure code without hindering the maintainability of
their projects. But there are still some concerns that need to be addressed
and that this study tries to bring awareness for:
